---
title: "PhyloMeth Invasion Project"
author: "M Mamantov"
date: "April 6, 2017"
output: html_document
---

Install all necessary packages.

```{r, eval=TRUE, include=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
#install.packages(c("ape","geiger", "OUwie", "TreeSim"))
library(ape)
library(geiger)
library(OUwie)
library(TreeSim)
library(phytools)
set.seed(1859)
```

Simulate a tree using a birth-death process where speciation rate is two and extinction rate is 0.5. Only include extant taxa.
```{r, eval=TRUE}
tree <- sim.bd.taxa(n=25,numbsim=1,lambda=2,mu=0.5,complete=FALSE)[[1]]
plot(tree)
```

Simulate a rate matrix based on invasion stats. There are four possible states: Naturalized native species with low plasticity (1), naturalized native with high plasticity (2), young invasive (3), and old invasive (4).

```{r, eval=TRUE}
rate.matrix <- matrix(data=c(-0.9,0.2,0,0,0.8,-.7,0,0.1,0.1,0.5,-0.5,0,0,0,0.5,-0.1), nrow=4, ncol=4)
rownames(rate.matrix) <- c("NNL","NNH","NI","OI")
colnames(rate.matrix) <- c("NNL","NNH","NI","OI")
rate.matrix
```

Use the tree to simulate data, using the `sim.char` function in `geiger`.First, I simulate continous plasticity data, which represents the absolute value of reaction norm slope. Then, I simulate data for invasion state using the rate matrix created above. 

```{r, eval=TRUE}
plasticity.data <- sim.char(tree,.2,model="BM",root=0.5)
invasion.data <- sim.char(tree, rate.matrix, model="discrete",root=1)
while(length(unique(invasion.data))!=4) {
  invasion.data <- sim.char(tree, rate.matrix, model="discrete",root=1)
}

full.data <- cbind.data.frame(invasion.data, plasticity.data)
colnames(full.data)<-c("Invasive Status","Plasticity")
full.data
```

Now, perform stochastic character mapping of invasive state.

```{r, eval=TRUE, message=FALSE}
traits.tree <- make.simmap(tree, x = invasion.data[,,1], model = "ARD", nsim=1, pi=c(.5,.5,0,0))
plotSimmap(traits.tree)
legend("bottomleft", legend=c("NNL","NNH","NI","NO"),lwd=1, bty="n", col=c("black","red","green","blue"))
```

Use OUwie to calculate the optimum trait values along each branch of the tree (thedas).

```{r, eval=TRUE}
OUwie.data <- cbind.data.frame(rownames(full.data),invasion.data,plasticity.data)
results <- OUwie(traits.tree,OUwie.data,model="OUM",simmap.tree=TRUE)
results
```

Is plasticity driving invasions?
```{r, eval=TRUE}
thetas <- results$theta[,1]
barplot(height=thetas, ylab= "Optimum Plasticity Value", names.arg= c("NNL","NNH","NI","OI"))
```
