
You can do this on your own data, or on included data here.


##Continuous data

```{r, eval=TRUE}
library(geiger)
library(picante)
tree.primates <- read.tree(text="((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") #using examples from ape ?pic
X <- c(4.09434, 3.61092, 2.37024, 2.02815, -1.46968)
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259)
names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago")
pic.X <- pic(X, tree.primates)
pic.Y <- pic(Y, tree.primates)

## lm(pic.X ~ pic.Y - 1) # regression through the origin
```

Now, positivitize the contrasts and do a regression through the origin.

##Discrete data

```{r, eval=TRUE}
require("corHMM")
?corHMM
data(primates)
print(primates)
require(phytools)
```


Just to make this a better dataset for our exercise, I'm going to change some of the states (I want to have all four trait combinations present).
For actual analyses, of course, DO NOT INVENT YOUR DATA.

First, a review of discrete state models:

```{r, eval=TRUE}
primates$trait[which(grepl("Hylobates",primates$trait[,1])),2]<-1

trait1<-primates$trait[,2]
names(trait1)<-primates$trait[,1]
plotSimmap(make.simmap(primates$tree, trait1), pts=FALSE, fsize=0.8)
rate.mat.er<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
print(rate.mat.er)
```

What does this matrix mean?
This matrix is showing the probability of transition from trait 0 to 1 and from 1 to 0. The rates are equal (both 1) because this is an equal rates model. 

```{r, eval=TRUE}
pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.er)
```

What do these results mean?
This matrix is giving us the maximum likelihood estimate of rates of evoltuion parameters. The rates of evolution are equal from traits 0 to 1 and from 1 to 0 because we are using an equal rates model. The matrix indicates taht the rate of evolution is 0.0107. 


```{r, eval=TRUE}
rate.mat.ard<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD")
print(rate.mat.ard)
```

And these?
This matrix now allows for different transition rates between states. The results mean that it is twice as likely to transition from state 0 to 1 than from state 1 to 0. 

```{r, eval=TRUE}
pp.ard<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
print(pp.ard)
```

which model is better?
```{r, eval=TRUE}
delta.AIC.pp.er <- pp.er$AIC - min(AIC.pp.er,AIC.pp.ard)
delta.AIC.pp.ard <- pp.ard$AIC - min(AIC.pp.er,AIC.pp.ard)
if (delta.AIC.pp.er == 0){
  print("Equal rates model is better")
} else{(print("All rates differ model is better"))}
```
The equal rates model has a lower AIC score. 


Now let's look at multiple traits.

This is a matrix with four states

```{r, eval=TRUE}
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state)
```

Convert the two binary traits into a single four character state

```{r, eval=TRUE}
fourstate.trait<-rep(NA,Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
	if(primates$trait[i,2]==0 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-0
	}	
	if(primates$trait[i,2]==0 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-1
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-2
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-3
	}	
}
fourstate.data<-data.frame(Genus_sp=primates$trait[,1], T1=fourstate.trait)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, model="ER", node.states="marginal"))
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=rate.mat.er.4state, node.states="marginal", model="ARD"))
rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")
print(rate.mat.ard.4state)
```

Now let's make the equivalent of a GTR matrix:

```{r, eval=TRUE}
rate.mat.gtr.4state<-rate.mat.ard.4state
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD"))
```


Now make a model like Pagel 1994

```{r, eval=TRUE}
print(rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-rate.par.drop(rate.mat.ard.4state, drop.par=c(3,5,8,10))
print
```

Now that you have some introduction, there are two routes:

##Route 1

**Construct a model to test if state 1 can never be lost**

```{r, eval=TRUE}
keep.1.matrix <- rate.mat.gtr.4state
keep.1.matrix[1,2:4]<-0

canbelost <- rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=rate.mat.gtr.4state, node.states="marginal", model="ARD")

cannotbelost <- rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=keep.1.matrix, node.states="marginal", model="ARD")

delta.AIC.canbelost <- canbelost$AIC - min(canbelost$AIC,cannotbelost$AIC)
delta.AIC.cannotbelost <- cannotbelost$AIC - min(canbelost$AIC,cannotbelost$AIC)
if (delta.AIC.canbelost == 0){
  print("Allowing transitions away from state 1 is a better model, indicating that state 1 can be lost")
} else{(print("Not allowing transitions away from state 1 is a better model, indicating that state 1 can never be lost"))}
```

**Experiment with the effects of frequencies at the root.**

**Create and use a model to see if transitions from 00 go to 11 only via 01.**

##Route 2

Maddison and FitzJohn (2015) pretty convincingly show (to me) that Pagel (1994) is just not a good method. Ok. So work on a fix. They point to [Read and Nee (1995)](http://dx.doi.org/10.1006/jtbi.1995.0047) as a low power but possible solution. Look at their appendix, especially, and write an implementation.
