
You can do this on your own data, or on included data here.


##Continuous data

```{r, eval=TRUE}
library(geiger)
library(picante)
tree.primates <- read.tree(text="((((Homo:0.21,Pongo:0.21):0.28,Macaca:0.49):0.13,Ateles:0.62):0.38,Galago:1.00);") #using examples from ape ?pic
X <- c(4.09434, 3.61092, 2.37024, 2.02815, -1.46968)
Y <- c(4.74493, 3.33220, 3.36730, 2.89037, 2.30259)
names(X) <- names(Y) <- c("Homo", "Pongo", "Macaca", "Ateles", "Galago")
pic.X <- pic(X, tree.primates)
pic.Y <- pic(Y, tree.primates)
```

Now, positivitize the contrasts and do a regression through the origin.
```{r,eval=TRUE}
for (i in sequence(length(X))) {
  if (X[i] < 0) {
    X[i] <- X[i]*-1
    Y[i] <- Y[i]*-1
  }
}
```
##Discrete data

```{r, eval=TRUE}
require("corHMM")
?corHMM
data(primates)
ls(primates)
print(primates)
require(phytools)
```


Just to make this a better dataset for our exercise, I'm going to change some of the states (I want to have all four trait combinations present).
For actual analyses, of course, DO NOT INVENT YOUR DATA.

First, a review of discrete state models:

```{r, eval=TRUE}
primates$trait[which(grepl("Hylobates",primates$trait[,1])),2]<-1

trait1<-primates$trait[,2]
names(trait1)<-primates$trait[,1]
plotSimmap(make.simmap(primates$tree, trait1), pts=FALSE, fsize=0.8)
rate.mat.er<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ER")
print(rate.mat.er)
```

What does this matrix mean?
#It shows that the rate of change from state 1 to 2 is equal to the rate of change from state 2 to 1.
```{r, eval=TRUE}
pp.er<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.er,node.states="marginal")
print(pp.er)
```

What do these results mean?

#The log likelihood is incredibly small, so the probability of the data is quite tiny.

```{r, eval=TRUE}
rate.mat.ard<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=2, model="ARD")
print(rate.mat.ard)
```

And these?

#This indicates that the transition rate from state 1 to 2 is twice as high as the transition rate from 2 to 1.

```{r, eval=TRUE}
pp.ard<-corHMM(primates$tree,primates$trait[,c(1,2)],rate.cat=1,rate.mat=rate.mat.ard,node.states="marginal")
print(pp.ard)
```

which model is better?

Now let's look at multiple traits.

This is a matrix with four states

```{r, eval=TRUE}
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rate.mat.er.4state)
```

Convert the two binary traits into a single four character state

```{r, eval=TRUE}
fourstate.trait<-rep(NA,Ntip(primates$tree))
for(i in sequence(Ntip(primates$tree))) {
	if(primates$trait[i,2]==0 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-0
	}	
	if(primates$trait[i,2]==0 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-1
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==0) {
		fourstate.trait[i]<-2
	}	
	if(primates$trait[i,2]==1 && primates$trait[i,3]==1) {
		fourstate.trait[i]<-3
	}	
}
fourstate.data<-data.frame(Genus_sp=primates$trait[,1], T1=fourstate.trait)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, model="ER", node.states="marginal"))
rate.mat.er.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ER")
print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat=rate.mat.er.4state, node.states="marginal", model="ARD"))
rate.mat.ard.4state<-rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=1, nstates=4, model="ARD")
print(rate.mat.ard.4state)
```

Now let's make the equivalent of a GTR matrix:

```{r, eval=TRUE}
rate.mat.gtr.4state<-rate.mat.ard.4state
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(1,4))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(2,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(3,8))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(4,6))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(5,7))
rate.mat.gtr.4state<-rate.par.eq(rate.mat.gtr.4state, c(6,7))
print(rate.mat.gtr.4state)

print(rayDISC(primates$tree, fourstate.data, ntraits=1, rate.mat= rate.mat.gtr.4state, node.states="marginal", model="ARD"))
```


Now make a model like Pagel 1994

```{r, eval=TRUE}
print(rate.mat.maker(rate.cat=1, hrm=FALSE, ntraits=2, nstates=2, model="ARD"))
rate.mat.pag94<-rate.par.drop(rate.mat.ard.4state, drop.par=c(3,5,8,10))
print(rate.mat.pag94)
```

Now that you have some introduction, there are two routes:

##Route 1

**Construct a model to test if state 1 can never be lost**
#Use rate.mat.maker, specifically drop.par to force a transition.
rate.mat.index=matrix(data = rate.mat.ard,nrow = 2, ncol = 2)
State1test<-rate.mat.maker(rate.cat=2, hrm=FALSE, ntraits=1, nstates=2, model="ARD")


**Experiment with the effects of frequencies at the root.**

**Create and use a model to see if transitions from 00 go to 11 only via 01.**

##Route 2

Maddison and FitzJohn (2015) pretty convincingly show (to me) that Pagel (1994) is just not a good method. Ok. So work on a fix. They point to [Read and Nee (1995)](http://dx.doi.org/10.1006/jtbi.1995.0047) as a low power but possible solution. Look at their appendix, especially, and write an implementation.
