---
title: "DiscreteCharacters"
author: "Brian O'Meara"
date: "2/16/2017"
output: html_document
---

For this exercise, explore looking at discrete character models. Note: they are not "discreet" character models -- they're actually pretty noisy.

Remember for homework, 

`git remote -v` gives list of remotes

`git fetch upstream`

`git pull upstream master` pulls from upstream (me).

change `eval=TRUE` for all the R blocks, debug, commit, and do a pull request.

These are useful packages in this area (far from exhaustive list).

```{r, eval=TRUE}
library(ape)
library(geiger)
library(phytools)
library(phangorn)
library(corHMM)
```

You'll need to get data into R in some way. 

```{r, eval=TRUE}
## tree

tree.nex = read.nexus("D:/EEB 603/Trees/consensusTree_10kTrees_Primates_Version3.nex")
phy <- multi2di(tree.nex, random=FALSE)


##  Discrete data

setwd("D:/EEB 603/Code/Run 1")
source("Macaque Project Functions.R")
source("Macaque Project Data Prep.R")

discrete.data.raw <- read.csv("D://EEB 603/KUPRI Data/KUPRI_MAST_WITH_DISC.csv", stringsAsFactors = FALSE)
discrete.data.prep <- as.matrix(discrete.data.raw[,6])
discrete.data.prep <-`rownames<-`(discrete.data.prep, c("Pan_troglodytes_troglodytes", "Gorilla_gorilla_gorilla", "Hylobates_lar", "Macaca_sylvanus","Macaca_thibetana", "Macaca_mulatta", "Macaca_fascicularis", "Macaca_fuscata", "Macaca_cyclopis", "Macaca_nemestrina", "Macaca_silenus", "Macaca_pagensis", "Macaca_nigra", "Macaca_maura", "Macaca_sinica", "Macaca_radiata", "Macaca_assamensis", "Macaca_arctoides","Papio_hamadryas", "Theropithecus_gelada", "Mandrillus_sphinx", "Cercocebus_torquatus", "Cercocebus_galeritus","Erythrocebus_patas","Cercopithecus_neglectus","Cercopithecus_campbelli","Cercopithecus_petaurista"))
discrete.factors <- factor(discrete.data.prep)
num.vec.disc <- as.matrix(as.numeric(discrete.factors))
num.vec.disc <-`rownames<-`(num.vec.disc, c("Pan_troglodytes_troglodytes", "Gorilla_gorilla_gorilla", "Hylobates_lar", "Macaca_sylvanus","Macaca_thibetana", "Macaca_mulatta", "Macaca_fascicularis", "Macaca_fuscata", "Macaca_cyclopis", "Macaca_nemestrina", "Macaca_silenus", "Macaca_pagensis", "Macaca_nigra", "Macaca_maura", "Macaca_sinica", "Macaca_radiata", "Macaca_assamensis", "Macaca_arctoides","Papio_hamadryas", "Theropithecus_gelada", "Mandrillus_sphinx", "Cercocebus_torquatus", "Cercocebus_galeritus","Erythrocebus_patas","Cercopithecus_neglectus","Cercopithecus_campbelli","Cercopithecus_petaurista"))

  pruned.phy <-name.check(phy, num.vec.disc)
  phy.discrete <- drop.tip(phy, pruned.phy$tree_not_data)


```

Data are often not right in some way. They might not match the taxa in your tree, there may be missing data, etc. `geiger::treedata` is a great function for getting a tree and data that match, but your data may need other cleaning. Do it as a function so it's repeatable.

```{r, eval=TRUE}
CleanData <- function(phy, data) {
	#treedata() in Geiger is probably my favorite function in R.

  #  Prep objects


  
  phy.reo <- ape::reorder.phylo(phy)
  data.reo <- geiger::treedata(phy.reo, data, sort=TRUE)$data
  

return(data.reo)
  }

# Now write the code to use CleanData() to actually clean your data

cleaned.discrete <- CleanData(phy.discrete, num.vec.disc)

```

It's critically important to LOOK at what you have. Are there weird values? Has the match between taxa and state gone correctly? Do you think you have binary data, but there's actually only state 1? Especially as data sets grow (yay), and are assembled using scripts rather than by error-prone, non-reproducable hands (double yay), scientists are increasingly less likely to deeply look at our data. That's bad -- don't be that person.

```{r, eval=TRUE}
VisualizeData <- function(data, main.char) {

  
  
  ##phenogram(phy, data, cex = .85, edge.width = .5, spread.labels = TRUE, fsize = .45)
   plot(density(data), main = main.char)
}

# disc.rownames <- rownames(cleaned.discrete)
# cleaned.discrete.df <- as.data.frame(cleaned.discrete, row.names = disc.rownames, stringsAsFactors = FALSE)

VisualizeData(cleaned.discrete, "cleaned.discrete")
```

First, let's use parsimony to look at ancestral states:

```{r, eval=FALSE}

##  Brian - my data arent genotypes but factors referencing foraging behaviors in primates, so this part got a bit muddled for me



cleaned.discrete.phyDat <- phangorn::phyDat(cleaned.discrete, type="______________") #phyDat is a data format used by phangorn
anc.p <- phangorn::ancestral.pars(tree, cleaned.discrete.phyDat)
plotAnc(tree, anc.p, 1)
```

Do you see uncertainty? What does it mean?

Now, plot the likelihood estimates. 

```{r, eval=FALSE}
anc.ml <- ancestral.pml(pml(tree, cleaned.discrete.phyDat), type="ml")
plotAnc(tree, anc.ml, 1)
```

How does this differ from parsimony? 

Why does it differ from parsimony?

What does uncertainty mean?

Now, to the biological questions. For many of these, `corHMM` will be a useful package.

1. How can you estimate transition rates between states? Do it.
2. How could you examine if transition rates are equal?
3. Think about the Lewis (2001) MKV model. Are your traits all variable? Will using this make sense for your data? Try using it. Do results change?
4. How could you test order of state evolution?