---
title: "DiscreteCharacters"
author: "Brian O'Meara"
date: "2/16/2017"
output: html_document
---

For this exercise, explore looking at discrete character models.

Remember for homework, change `eval=TRUE` for all the R blocks, debug, commit, and do a pull request.

These are useful packages in this area (far from exhaustive list).

```{r, eval=TRUE}
library(ape)
library(geiger)
library(phytools)
library(phangorn)
```

library(ape)

You'll need to get data into R in some way. 

```{r, eval=TRUE}
tree <- read.tree("Maggie_nexus.tre")
discrete.data <- read.delim(file="edriodata.csv",stringsAsFactors=FALSE)
ape::compute.brlen(tree)
tree<-ape::compute.brlen(tree)
```


Data are often not right in some way. They might not match the taxa in your tree, there may be missing data, etc. `geiger::treedata` is a great function for getting a tree and data that match, but your data may need other cleaning. Do it as a function so it's repeatable.

```{r, eval=TRUE}
CleanData <- function(tree, discrete.data) {
  tree$tip.label <- gsub("_", " ", tree$tip.label)
  rownames(discrete.data) <- discrete.data[,1]
  # make it a character matrix
  discrete.data <- as.matrix(discrete.data[,-1])
  # define ambiguous states "N" and "?"
  data <- phyDat(as.matrix(discrete.data), type="USER", levels=c("0", "1", "2"), ambiguity = c("N", "?"))
  # reorder data
  data <- subset(data, tree$tip.label) 
  list(phy=tree, data = data)
}

cleaned.results <- CleanData(tree, discrete.data)
cleaned.tree <- cleaned.results$phy
cleaned.data <- cleaned.results$data

match(cleaned.tree$tip.label, names(cleaned.data))  # should be 1:15
as.character(cleaned.data) # should look like the original

```

It's critically important to LOOK at what you have. Are there weird values? Has the match between taxa and state gone correctly? Do you think you have binary data, but there's actually only state 1? Especially as data sets grow (yay), and are assembled using scripts rather than by error-prone, non-reproducable hands (double yay), scientists are increasingly less likely to deeply look at our data. That's bad -- don't be that person.

```{r, eval=TRUE}
VisualizeData <- function(tree, data) {
  plot(tree)
  print(data)
}

VisualizeData(cleaned.tree, cleaned.data)

discrete.data.matrix <- as.matrix(cleaned.data)
rownames(discrete.data.matrix) <- rownames(cleaned.data)

print(discrete.data.matrix)
```

First, let's use parsimony to look at ancestral states:

```{r, eval=TRUE}
anc.p <- ancestral.pars(cleaned.tree, cleaned.data)
print(anc.p)
plotAnc(tree, anc.p, attr(anc.p, "index")[1])
plotAnc(tree, anc.p, attr(anc.p, "index")[25])
```

Do you see uncertainty? What does it mean?
The uncertainty here means that it is possible that there are different ways to achieve this result or different numbers of changes that could happen at those taxa.

Now, plot the likelihood estimates. 

```{r, eval=TRUE}
anc.ml <- ancestral.pml(pml(cleaned.tree, cleaned.data), type="ml")
plotAnc(tree, anc.ml, 1)
```

Why does it differ from parsimony?
Maximum likelihood differs from parsimony because parsimony is about finding the tree with the least amount of changes to get the end result whereas likelihood is about calculating the likelihood that this tree is correct based on the transitions from state to state. 

What does uncertainty mean?


Now, to the biological questions. For many of these, `corHMM` will be a useful package.

1. How can you estimate transition rates between states? Do it.

2. How could you examine if transition rates are equal?
3. Think about the Lewis (2001) MKV model. Are your traits all variable? Will using this make sense for your data? Try using it. Do results change?
4. How could you test order of state evolution?