---
title: 'PhyloMeth: Correlations'
author: "Brian O'Meara"
output: html_document
date: '`r format(Sys.time(), "Last updated %H:%M:%S %p %d %B, %Y")`'
---

*In progress*

##Objectives
By the end of this week, you will:

* Understand the importance of dealing with correlations in an evolutionary manner
* Know methods for looking at correlations of continuous and discrete traits
* Be able to point to reasons to be concerned.

Make sure to **read the relevant papers**: https://www.mendeley.com/groups/8111971/phylometh/papers/added/0/tag/week6/    

First, let's get a tree:

```{r, warning=FALSE, message=FALSE, fig.height=12}
library(rotl)
library(ape)
phy <- get_study_tree("ot_485", "tree1")
plot(phy, cex=0.5)
axisPhylo(backward=TRUE)
```

Note that this tree is a chronogram. 

Let's simulate data on this tree. But what model to use? For now, let's assume we are looking at continuous traits, things like body size. Over evolutionary time, these probably undergo a series of changes that then get added up. A species has an average mass of 15 kg, then it goes to 15.1 kg, then 14.8 kg, and so forth. But how could those changes be distributed?

Start with a uniform distribution. Take a starting value of 0, then pick a number from -1 to 1 to add to it (in other words, `runif(n=1, min=-1, max=1)`). There are efficient ways to do this for many generations, but let's do the obvious way: a simple `for` loop. Do it for 100 generations.

```{r}
ngen <- 100
positions <- c(0, rep(NA,ngen))
for (i in sequence(ngen)) {
  positions[i+1] <- positions[i] + runif(1,-1,1)
}
plot(x=positions, y=sequence(length(positions)), xlab="trait value", ylab="generation", bty="n", type="l")
```

We can repeat this simulation many times and see what the pattern looks like:

```{r}
ngen <- 100
nsims <- 500
final.positions <- rep(NA, nsims)
# make a plot to hold our lines
plot(x=c(-1,1)*ngen, y=c(1, 1+ngen), xlab="trait value", ylab="generation", bty="n", type="n") 
for (sim.index in sequence(nsims)) {
  positions <- c(0, rep(NA,ngen))
  for (i in sequence(ngen)) {
    positions[i+1] <- positions[i] + runif(1,-1,1)
  }
  lines(positions, sequence(length(positions)), col=rgb(0,0,0,0.1))
  final.positions[sim.index] <- positions[length(positions)]
}
```

Well, that may seem odd: we're adding a bunch of uniform random values between -1 and 1 (so, a flat distribution) and we get something that definitely has more lines ending up in the middle than further out. Look just at the distribution of final points:

```{r}
plot(density(final.positions), col="black", bty="n")
```

Which looks almost normal. Ok, let's try a weird distribution:

```{r}
rweird <- function() {
  displacement <- 0
  if(runif(1,-2,2) < .1) {
    displacement <- rnorm(1, 7, 3) + runif(1,0,7)
  } else {
    displacement <- 0.5 * rexp(1, 0.3) - 1 
  }
  displacement <- displacement + round(runif(1,1,100) %% 7)
  return(displacement)  
}

plot(density(replicate(100000, rweird())), bty="n")
```

When we ask `rweird()` for a number it sometimes gives us a normally distributed number multiplied by a unifor distribution, other times it gives us an exponentially distributed number, and then adds the remainder that comes when you divide a random number by 7. So, not exactly a simple distribution like uniform, normal, or Poisson. So, repeating the simulation above but using this funky distribution:

```{r}
ngen <- 100
nsims <- 500
final.positions <- rep(NA, nsims)
# make a plot to hold our lines
plot(x=c(-100,1200), y=c(1, 1+ngen), xlab="trait value", ylab="generation", bty="n", type="n") 
for (sim.index in sequence(nsims)) {
  positions <- c(0, rep(NA,ngen))
  for (i in sequence(ngen)) {
    positions[i+1] <- positions[i] + rweird()
  }
  lines(positions, sequence(length(positions)), col=rgb(0,0,0,0.1))
  final.positions[sim.index] <- positions[length(positions)]
}
```

And now let's look at final positions again:

```{r}
plot(density(final.positions), col="black", bty="n")
```

Again, it looks pretty much like a normal distribution. You can try with your own wacky distribution, and this will almost always happen (as long as the distribution has finite variance).
