---
title: "Continuous character models"
author: "Brian C. O'Meara"
date: "2/20/2017"
output: html_document
---

First get packages we need

```{r, eval=TRUE}
library(ape) #utility fns
library(geiger) #utilty fns
library(OUwie)
library("phangorn")
```

Now get the tree and data. For these exercises, knowing uncertainty in your measurements can also be important. (remember for homework to change `eval=FALSE` to `eval=TRUE`).

```{r eval=TRUE}

setwd("D:/EEB 603/Code/Run 1")
source('D:/EEB 603/Code/Run 1/ContinuousFunctions.R')
source("Macaque Project Data Prep.R")


# Read in tree

tree.nex = read.nexus("D:/EEB 603/Trees/consensusTree_10kTrees_Primates_Version3.nex")
phy <- multi2di(tree.nex, random=FALSE)
treefile1 = phy

##  Pull in morphometric/behavior data to be compiled

discrete.data.raw <- read.csv("D://EEB 603/KUPRI Data/KUPRI_MAST_WITH_DISC.csv", stringsAsFactors = FALSE)
discrete.data.prep <- as.matrix(discrete.data.raw[,6])
discrete.data.prep <-`rownames<-`(discrete.data.prep, c("Pan_troglodytes_troglodytes", "Gorilla_gorilla_gorilla", "Hylobates_lar", "Macaca_sylvanus","Macaca_thibetana", "Macaca_mulatta", "Macaca_fascicularis", "Macaca_fuscata", "Macaca_cyclopis", "Macaca_nemestrina", "Macaca_silenus", "Macaca_pagensis", "Macaca_nigra", "Macaca_maura", "Macaca_sinica", "Macaca_radiata", "Macaca_assamensis", "Macaca_arctoides","Papio_hamadryas", "Theropithecus_gelada", "Mandrillus_sphinx", "Cercocebus_torquatus", "Cercocebus_galeritus","Erythrocebus_patas","Cercopithecus_neglectus","Cercopithecus_campbelli","Cercopithecus_petaurista"))
discrete.factors <- factor(discrete.data.prep)
num.vec.disc <- as.matrix(as.numeric(discrete.factors))
num.vec.disc <-`rownames<-`(num.vec.disc, c("Pan_troglodytes_troglodytes", "Gorilla_gorilla_gorilla", "Hylobates_lar", "Macaca_sylvanus","Macaca_thibetana", "Macaca_mulatta", "Macaca_fascicularis", "Macaca_fuscata", "Macaca_cyclopis", "Macaca_nemestrina", "Macaca_silenus", "Macaca_pagensis", "Macaca_nigra", "Macaca_maura", "Macaca_sinica", "Macaca_radiata", "Macaca_assamensis", "Macaca_arctoides","Papio_hamadryas", "Theropithecus_gelada", "Mandrillus_sphinx", "Cercocebus_torquatus", "Cercocebus_galeritus","Erythrocebus_patas","Cercopithecus_neglectus","Cercopithecus_campbelli","Cercopithecus_petaurista"))


## Force trees to be ultrametric

  ##  A function to do this 

  force.ultrametric<-function(tree,method=c("nnls","extend")){
    method<-method[1]
    if(method=="nnls") tree<-nnls.tree(cophenetic(tree),tree,
                                       rooted=TRUE,trace=0)
    else if(method=="extend"){
      h<-diag(vcv(tree))
      d<-max(h)-h
      ii<-sapply(1:Ntip(tree),function(x,y) which(y==x),
                 y=tree$edge[,2])
      tree$edge.length[ii]<-tree$edge.length[ii]+d
    } else 
      cat("method not recognized: returning input tree\n\n")
    tree
  }

 
 # discrete 

  pruned.phy <-name.check(treefile1, num.vec.disc)
  phy.discrete <- drop.tip(treefile1, pruned.phy$tree_not_data)
  branch.time <- branching.times(phy.discrete)
  ult.phy.disc <- force.ultrametric(phy.discrete)
 
  is.ultrametric(ult.phy.disc)
  
  # continous
  
  continuous.data <- log(phenotype.z.complete[,1])
  pruned.phy.cont <- name.check(treefile1, continuous.data)
  phy.cont <- drop.tip(treefile1, pruned.phy.cont$tree_not_data)
  branch.time.cont <- branching.times(phy.cont)
  ult.phy.cont <- force.ultrametric(phy.cont)
  
  is.ultrametric(ult.phy.cont)


```

A function to clean data, make sure taxon names match between tree and data, etc. 

```{r, eval=TRUE}
CleanData <- function(tree.phy, data.phy, col.no) {
  
  phy <- ape::reorder.phylo(tree.phy)
  data1 <- geiger::treedata(phy, data.phy, sort=TRUE)$data
  
    ##  Since we're doing univar for now, I'm asking for the vector of pheno vals I need
  
   return(data1[,col.no])
  
}
```

# A function to plot data. Look at `phytools::contMap()`. This is all part of checking: do your data all seem sensible? **LOOK AT IT**.

```{r, eval=TRUE}

  pruned.phy.cont <- name.check(treefile1, continuous.data)
  phy.cont <- drop.tip(treefile1, pruned.phy.cont$tree_not_data)
  branch.time.cont <- branching.times(phy.cont)
  ult.phy.cont <- force.ultrametric(phy.cont)
 

Plot.phy <- function(tree.phy, main.char){
  
  plot.phy <- plot(tree.phy, main = main.char, show.tip.label = FALSE)
  
  return(plot.phy)
  
}

VisualizeData <- function(phytree, dat.mat, main.char){

    ## make a phenogram

    phytools::phenogram(phytree, dat.mat, main = main.char, cex = .85, edge.width = .5, spread.labels = TRUE, fsize = .45)


  ##  Plot density
    plot(density(dat.mat), main = main.char)
    
}

## Clean the data 
  
cleaned.continuous <- CleanData(ult.phy.cont, continuous.data, 1)
cleaned.discrete <- CleanData.discrete(ult.phy.disc, num.vec.disc, 1)
VisualizeData(ult.phy.cont, cleaned.continuous, "Continuous_IOB")
VisualizeData(ult.phy.disc, cleaned.discrete, "FORAGE_POOLED")

```





First, start basic. What is the rate of evolution of your trait on the tree? 

```{r, eval=TRUE}
#First, start basic. What is the rate of evolution of your trait on the tree? 

BM1 <- fitContinuous(ult.phy.cont, cleaned.continuous, model="BM")
print(paste("The rate of evolution is", "0.006838", "in units of", "sigma^2/time"))
```

Important: What are the rates of evolution? In what units?


```{r, eval=TRUE}
OU1 <- fitContinuous(ult.phy.cont, cleaned.continuous, model="OU")
par(mfcol=(c(1,2)))
mod.alpha = OU1$opt$alpha
Plot.phy(ult.phy.cont, "IOB_ALL")
ou.tree <- geiger::rescale(ult.phy.cont, model="OU", alpha = mod.alpha)
Plot.phy(ou.tree, "IOB_ALL_RESCALE")
```

How are the trees different?

Compare trees

```{r, eval=TRUE}
AIC.BM1 <- BM1$opt$aic
AIC.OU1 <- OU1$opt$aic
min.AIC = min(AIC.BM1, AIC.OU1)
delta.AIC.BM1 <- BM1$opt$aic - min.AIC
delta.AIC.OU1 <- OU1$opt$aic - min.AIC
print(c("Delta AIC BM1:", delta.AIC.BM1))
print(c("Delta AIC OU1:", delta.AIC.OU1))
```


##OUwie runs##

This takes longer than you may be used to. 

We're a bit obsessive about doing multiple starts and in general performing a thorough numerical search. It took you 3+ years to get the data, may as well take an extra five minutes to get an accurate answer

First, we need to assign regimes. The way we do this is with ancestral state estimation of a discrete trait. We can do this using ace() in ape, or similar functions in corHMM or diversitree. Use only one discrete char.

```{r, eval=TRUE}

## get discrete species into df
disc.species <- rownames(num.vec.disc)
disc.traits.conv <- as.matrix(cleaned.discrete)
disc.traits <- `row.names<-.default`(disc.traits.conv, value = NULL)
disc.df <- as.data.frame(disc.traits, row.names = disc.species, stringsAsFactors = FALSE)
disc.phylo <- as.data.frame(disc.traits.conv[match(ult.phy.disc$tip.label,rownames(disc.traits.conv)),])
disc.phylo.vector <- disc.phylo[,1]
names(disc.phylo.vector) <- rownames(disc.phylo)

ult.phy.disc.tinybrlen <- ult.phy.disc
ult.phy.disc.tinybrlen$edge.length[which(ult.phy.disc.tinybrlen$edge.length==0)] <- 1e-08

reconstruction.info <- ape::ace(disc.phylo.vector,ult.phy.disc.tinybrlen, type="discrete", method="ML", CI=FALSE, model = "ER")

# best.states <- colnames(reconstruction.info)[apply(reconstruction.info$call, 1, which.max)]

```

Now add these labels to your tree.

```{r, eval=FALSE}

##  Get continuous data into df
cont.species <- rownames(as.matrix(continuous.data))
cont.traits.conv <- as.matrix(cleaned.continuous)
cont.traits <- `row.names<-.default`(cont.traits.conv, value = NULL)
cont.df <- as.data.frame(cont.traits, row.names = cont.species, stringsAsFactors = FALSE)
cont.phylo <- as.data.frame(cont.traits.conv[match(ult.phy.cont$tip.label,rownames(cont.traits.conv)),])


# labeled.tree <- ________________

ult.phy.cont.tinybrlen <- ult.phy.cont
ult.phy.cont.tinybrlen$edge.length[which(ult.phy.cont.tinybrlen$edge.length==0)] <- 1e-08

nodeBased.OUMV <- OUwie(ult.phy.cont.tinybrlen, cont.phylo, model="OUMV", simmap.tree=FALSE, diagn=FALSE)
print(nodeBased.OUMV)

##Error in `[.data.frame`(data, , 3) : undefined columns selected


```

What do the numbers mean?

Now run all OUwie models:

```{r, eval=FALSE}
models <- c("BM1","BMS","OU1","OUM","OUMV","OUMA","OUMVA")
results <- lapply(models, RunSingleOUwieModel, phy=ult.phy.cont.tinybrlen, data=cont.phylo)

AICc.values<-sapply(results, "[[", "AICc")
names(AICc.values)<-models
AICc.values<-AICc.values-min(AICc.values)


print(AICc.values) #The best model is the one with smallest AICc score

best<-results[[which.min(AICc.values)]] #store for later

print(best) #prints info on best model
```

We get SE for the optima (see nodeBased.OUMV$theta) but not for the other parameters. Let's see how hard they are to estimate. 
First, look at ?OUwie.fixed to see how to calculate likelihood at a single point.

```{r, eval=FALSE}
?OUwie.fixed
```

Next, keep all parameters but alpha at their maximum likelihood estimates (better would be to fix just alpha and let the others optimize given this constraint, but this is harder to program for this class). Try a range of alpha values and plot the likelihood against this.

```{r, eval=FALSE}
alpha.values<-seq(from= _______________ , to= _______________ , length.out=50)
```

Keep it simple (and slow) and do a for loop:

```{r, eval=FALSE}
likelihood.values <- rep(NA, length(alpha.values))
for (iteration in sequence(length(alpha.values))) {
	likelihood.values[iteration] <- OUwie.fixed(tree, trait, model="OUMV", alpha=rep(alpha.values[iteration],2), sigma.sq=best$solution[2,], theta=best$theta[,1])$loglik
}

plot(x= _______________ , y= _______________, xlab="_______________", ylab="_______________", type="l", bty="n")
points(x=best$solution[1,1], y=best$loglik, pch=16, col="red")
text(x=best$solution[1,1], y=best$loglik, "unconstrained best", pos=4, col="red")
```

A rule of thumb for confidence for likelihood is all points two log likelihood units worse than the best value. Draw a dotted line on the plot to show this

```{r, eval=FALSE}
abline(h=_______________, lty="dotted") #Two log-likelihood 
```

Now, let's try looking at both theta parameters at once, keeping the other parameters at their MLEs

```{r, eval=FALSE}
require("akima")
nreps<-400
theta1.points<-c(best$theta[1,1], rnorm(nreps-1, best$theta[1,1], 5*best$theta[1,2])) #center on optimal value, have extra variance
theta2.points<-c(best$theta[2,1], rnorm(nreps-1, best$theta[2,1], 5*best$theta[2,2])) #center on optimal value, have extra variance
likelihood.values<-rep(NA,nreps)

for (iteration in sequence(nreps)) {
	likelihood.values[iteration] <- OUwie.fixed(tree, trait, model="OUMV", alpha=best$solution[1,], sigma.sq=best$solution[2,], theta=c(theta1.points[iteration], theta2.points[iteration]))$loglik
}
```


Think of how long that took to do 400 iterations. Now remember how long the search took (longer).

```{r, eval=FALSE}
likelihood.differences<-(-(likelihood.values-max(likelihood.values)))
```

We are interpolating here: contour wants a nice grid. But by centering our simulations on the MLE values, we made sure to sample most thoroughly there

```{r, eval=FALSE}
interpolated.points<-interp(x=theta1.points, y=theta2.points, z= likelihood.differences, linear=FALSE, extrap=TRUE, xo=seq(min(theta1.points), max(theta1.points), length = 400), yo=seq(min(theta2.points), max(theta2.points), length = 400))
	
contour(interpolated.points, xlim=range(c(theta1.points, theta2.points)),ylim=range(c(theta1.points, theta2.points)), xlab="Theta 1", ylab="Theta 2", levels=c(2,5,10),add=FALSE,lwd=1, bty="n", asp=1)

points(x=best$theta[1,1], y=best$theta[2,1], col="red", pch=16)

points(x=trait$X[which(trait$Reg==1)],y=rep(min(c(theta1.points, theta2.points)), length(which(trait$Reg==1))), pch=18, col=rgb(0,0,0,.3)) #the tip values in regime 1, plotted along x axis
points(y=trait$X[which(trait$Reg==2)],x=rep(min(c(theta1.points, theta2.points)), length(which(trait$Reg==2))), pch=18, col=rgb(0,0,0,.3)) #the tip values in regime 2, plotted along y axis
```

The below only works if the discrete trait rate is low, so you have a good chance of estimating where the state is. If it evolves quickly, hard to estimate where the regimes are, so some in regime 1 are incorrectly mapped in regime 2 vice versa. This makes the models more similar than they should be. See Revell 2013, DOI:10.1093/sysbio/sys084 for an exploration of this effect.

```{r, eval=FALSE}
library(phytools)
trait.ordered<-data.frame(trait[,2], trait[,2],row.names=trait[,1])
trait.ordered<- trait.ordered[tree$tip.label,]
z<-trait.ordered[,1]
names(z)<-rownames(trait.ordered)
tree.mapped<-make.simmap(tree,z,model="ER",nsim=1)
leg<-c("black","red")
names(leg)<-c(1,2)
plotSimmap(tree.mapped,leg,pts=FALSE,ftype="off", lwd=1)

simmapBased<-OUwie(tree.mapped,trait,model="OUMV", simmap.tree=TRUE, diagn=FALSE)
print(simmapBased)
print(best)
```

How does this compare to our best model from above? Should they be directly comparable?

